Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-07-11 Manuscript/manuscript.Rmd
intro edits

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed the spatial organization of the human genome  [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Given their high degree of conservation between cell types and species, it has been proposed that TADs represent the fundamental building blocks of organization, giving the genome its 3D structure [@rocha2015breaking]. Studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome. It has been thought that the collection of particular functional genomic elements may contribute to the formation of TADs. Namely, the CCCTC-binding factor (CTCF) has been shown to be enriched at TAD boundaries [@ghirlando2016ctcf]. Additional transcription factors including SMC1, SMC3, and RAD21 have also been shown to colocalize near TAD boundaries [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, chromatin has been shown to exhibit certain modifications at TAD boundaries, evidenced by the frequent observation of the H3K36me3 histone mark [@dixon2012topological].  However, the growing number of publically available annotated genomic elements and there collinearity pose a significant problem for determining the most important epigenomic factors that contribute to TAD boudandary formation.

@@

Distinct patterns in epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries [@dixon2015chromatin; @di2017novo]. Currently, classification models have been used to draw associations between these functional annotations and genomic regions that include TAD boundaries.

@@

##################
Recent research has identified key functional genomic annotations enriched at boundaries. Namely, the CCCTC-binding factor  (CTCF), has been shown to be associated with TAD boundaries [@ghirlando2016ctcf]. Other transcription factors including SMC1, SMC3, and RAD21 have also been shown to colocalize near TAD boundaries [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Together, these components are hypothesized to be what causes loops in the DNA, contributing to higher order chromatin organization. Additional binding proteins, known to mediate enhancer-promoter interactions including YY1 and ZNF143, have been shown to influence the delineation of topological domains [@weintraub2017yy1; @gonzalez2017znf143; @heidari2014genome]. <!--However, there is a growing number of publically available annotated genomic features, and it is unclear if current methods are robust enough to identify all of them. [??? This is very weak and undefined. You started with mentioning prediction - finish with unsolved problems with it. !!!editted]--> [??? What's the summary/conclusion/problem?]

Current methods used to predict TAD boundaries rely on the use of classification models to draw associations between functional annotations and genomic regions that include TAD boundaries. However, many key characteristics of 3D genomic data remain unexplored, potentially limiting performance of predictive modeling. Specifically, the minority of TAD boundaries as compared with the rest of the genome creates proportionally small TAD boundary regions versus non-TAD boundary regions. These TAD boundary regions make up the minority class, while the non-TAD boundary regions make up the remaining majority class. The large disparity is sizes between minority and majority classes creates imbalanced data, as can be characterized by the *proportion of imbalance* (PI). Here, PI is defined as the ratio of the set minority class observations to the sum of both sets of minority and majority class observations. A dataset is considered imbalanced if PI<0.5. [??? Not really. Poor metric choice.] <!--Specifically, the minority of TAD boundaries as compared with the rest of the genome creates heavily imbalanced classes, characterized by ratios of positive/negative classes used to train the models. Here, the imbalance ratio is defined as the ratio of majority classes (regions not including TAD boundaries) to minority classes (regions including TAD boundaries) [@Orriols-Puig2008].--> Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Furthermore, the use of different levels of association between TAD boundaries and genomic features (number, or proportion of overlaps, or distance) as predictors remain underexplored. Lastly, it remains unclear how increased resolution of Hi-C data might affect the performance of predictive models. 
##################

Recently, an L1-regularized multiple logistic regression model, called *HiCfeat*, was developed to predict TAD boundaries using transcription factor binding sites on GM12878 [@mourad2016computational]. Based on a 1 kb binned genomic data matrix and 20 kb flanked TAD boundaries, the model was trained using proportions of overlaps between ChiP-seq peak defined functional element coordinates and genomic bins. However, several problems exist with this study. First, many transcription factors are known to colocalize, and therefore exhibit high correlations. Given a group of features that are highly correlated, L1-regularization tends to pick only one of them, while shrinking the others to zero [@zou2005regularization]. Here, ensemble learning techniques could offer a more robust method toward identifying  annotations associated with TAD boudaries. Likewise, it is unclear how other sets of epigenomic features affect prediction, including chromatin state segmentations and histone modifications. Secondly, the class imbalance issue was not addressed prior to modelling. Instead, model performance was evaluated using ROC (receiver operating characteristic) curves and subsequent AUC (area under the curve). It has been shown that AUC is insensitive to class imbalance, creating artificially inflated values inlfuenced by the majority class [@prati2011survey; @wei2013role]. It is unclear how simple data level solutions to class imbalance, such random under-sampling and random over-sampling affect TAD boundary prediction. Likewise, a more composite metric such as Matthews Correlation Coefficient (MCC), might provide more sensitive assessment of model performance. Thirdly, the relationship between annotations and boundaries was quantified using proportions of overlaps. This type of predictor ignores distances, which offers a much more spatially descriptive feature space and could improve predictive performance. Lastly, it is unclear how *HiCfeat* would perform on other cell lines and Hi-C data resolutions, giving *HiCfeat* a limited application in TAD boundary prediction.

<!--[Paragraph about disconnect between Hi-C and epigenomic data]--> 
Despite the success of Hi-C technologies in revealing the high-order levels of genome organization (chromosome territories, A/B compartments, TADs, sub-TADs, chromatin loops), resolution of Hi-C data (~750kb [??? REF, Bonev 2017?]) remains well below resolution of epigenomic assays (hundredths of bases for ChIP-seq). One of the key missing pieces to connect 3D genome architecture to transcription has been the inability of Hi-C to attain the resolution necessary to discern gene-level features of chromatin. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation of gene expression and its potential link to chromatin architecture. Although computational [??? REFS?](Zhang et al., “HiCPlus.”, Carron et al., “Boost-HiC.”) and technological [@Hsieh:2019aa][@Ma:1998aa][@Belaghzal2017] methods for enhancing resolution of Hi-C data started to appear, they [??? remains to write about their pitfalls]. One aim of this study is to predict the location of TAD boundaries on a scale comparable to epigenomic assays. For this, we developed a model trained on a relatively coarse Hi-C data for predicting 3D features (TAD boundaries) from 1D data (epigenomic datasets).

>>>>

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed the spatial organization of the human genome  [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Given their high degree of conservation between cell types and species, it has been proposed that TADs represent the fundamental building blocks of organization, giving the genome its 3D structure [@rocha2015breaking]. Studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome. It has been thought that the collection of particular functional genomic elements may contribute to the formation of TADs. Namely, the CCCTC-binding factor (CTCF) has been shown to be enriched at TAD boundaries [@ghirlando2016ctcf]. Additional transcription factors including SMC1, SMC3, and RAD21 have also been shown to colocalize near TAD boundaries [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, chromatin has been shown to exhibit certain modifications at TAD boundaries, evidenced by the frequent observation of the H3K36me3 histone mark [@dixon2012topological].  However, the growing number of publically available annotated genomic elements and there collinearity pose a significant problem for establishing methods robust enough to determine the most important epigenomic factors that contribute to TAD boudandary formation.

The distinct patterns in epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches toward TAD boundary prediction  <!--[@dixon2015chromatin; @di2017novo]-->. Currently, classification models have been used to draw associations between these functional annotations and genomic regions that include TAD boundaries. However, many key characteristics of 3D genomic data remain unexplored, potentially limiting the performance of predictive modeling.

First, the use of different levels of association between TAD boundary regions and genomic features as predictors remain underexplored. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association as it simply counts the total number of peaks defining a genomic annotation that are inside a given TAD boundary region. Percents of overlaps (**overlap percent**, OP) represents a more precise measure of association as it considers the proportion of overlap betwee a genomic annotation and a boundary region. However, the distance between boundary regions and genomic annotations are currently unexplored. This association allows for quantifying the spatial relationship between genomic elements and boundary regions. We hypothesize that the inclusion of this type of association will correspond with more accurate identification of functional genomic elements association with TAD boundary formation.

Second, the minority of TAD boundaries as compared with the rest of the genome creates proportionally small TAD boundary regions versus non-TAD boundary regions. These boundary regions make up the minority class, while the non-boundary regions make up the remaining majority class. The large disparity is sizes between minority and majority classes creates imbalanced data, as can be characterized by the *proportion of imbalance* (PI). Here, PI is defined as twice the ratio of the set minority class observations to the sum of both sets of minority and majority class observations. That is, the PI is bounded by 0 and 1. A dataset is considered balanced if PI$\approx$1, and imbalanced if PI$<<$1. Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. We hypothesize that the implementation of novel data resampling techniques will significantly improve model performance. Futhermore, the choice of performance metric can mask the poor performance of models implemented on imbalanced data.  It has been shown that conventional metrics such as AUC and accuracy are insensitive to class imbalance, creating artificially inflated values influenced by the majority class [@prati2011survey; @wei2013role]. Therefore, we hypothesize that a more composite metric such as Matthews Correlation Coefficient (MCC), might provide more sensitive assessment of model performance, given that it penalizes for both false positives and false negatives.

Thirdly, TAD boundary location is known to be heavily influenced by the resolution of the subsequent Hi-C interaction matrix [@sexton2012three]. Higher resolutions require a larger number of mapped reads, thereby creating larger data matrices, increasing computation time. However, the increased resolution available in newer datasets has highlighted the existence of a hierarchical structure of TADs inside other TADs [@pal2019hi]. It is unclear how varying levels of resolution affect the association between functional genomic annotations and TAD boundary regions.
<!--[link these two paragraphs]-->
<!--[Paragraph about disconnect between Hi-C and epigenomic data] 
Despite the success of Hi-C technologies in revealing the high-order levels of genome organization (chromosome territories, A/B compartments, TADs, sub-TADs, chromatin loops), resolution of Hi-C data (~750kb [??? REF, Bonev 2017?]) remains well below resolution of epigenomic assays (hundredths of bases for ChIP-seq). One of the key missing pieces to connect 3D genome architecture to transcription has been the inability of Hi-C to attain the resolution necessary to discern gene-level features of chromatin. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation of gene expression and its potential link to chromatin architecture. Although computational [??? REFS?](Zhang et al., “HiCPlus.”, Carron et al., “Boost-HiC.”) and technological [@Hsieh:2019aa][@Ma:1998aa][@Belaghzal2017] methods for enhancing resolution of Hi-C data started to appear, they [??? remains to write about their pitfalls]. One aim of this study is to predict the location of TAD boundaries on a scale comparable to epigenomic assays. For this, we developed a model trained on a relatively coarse Hi-C data for predicting 3D features (TAD boundaries) from 1D data (epigenomic datasets).-->

@@

Recently, an L1-regularized multiple logistic regression model, called *HiCfeat*, was developed to predict TAD boundary regions using a collection of 69 transcription factor binding sites on the GM12878 cell line [@mourad2016computational].

Another method used an L2-based regularized linear model to discriminate between what was termed as "boundary-" and "TAD-segments" of the hESC cell line [@hong2017computational].

A method devolped in 2015 used a tree based model to predict TAD boundary regions using 9 different cell line specific histone modifications, as well as consensus CTCF peaks, across 8 cell lines [@huang2015predicting].

@@
